The Design and Implementation of the Synchronous Language \CEU

### 2016-09-24 paper.tex
(*) TODO: data dependency, constructive semantics, internal events
(>) more neutral style

Despite its success and influence, Esterel has an overly complex semantics that 
requires careful static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems%
%These problems have an extensive coverage in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
The complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Above all, it also affects the programmer's understanding about the code, 
who, ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of loose and 
non-deterministic execution for intra-reaction statements.
Non determinism prevents threads from interacting with stateful system calls safely 
and makes shared-memory concurrency not as straightforward as reading and 
writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU has a simple semantics with fine-grained control for intra-reaction 
execution, and is amenable to concurrency checks that improve safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static concurrency checks to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.checks}).
%
\item Safe integration with \C that enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

@@

Note that such analysis is only possible due to the uniqueness of input events 
within reactions.
Otherwise, any two trail segments in parallel could be concurrent, even if they 
react to different input events.
%
Note also that the static checks are optional and do not affect the semantics
of the program.
% TODO-no: static analysis also simpler
% highlight that is not only important to have a comptutally-sound algorithm, 
% but that something that programmers can understand

@@

In \CEU, any identifier prefixed with an underscore is passed unchanged to 
the \C compiler that generates the final binary.
Therefore, access to \C is seamless and, more importantly, easily trackable.
%
Similarly to Esterel with the \code{call} primitive, external calls are assumed 
to be instantaneous~\cite{esterel.primer}.
%
This way, programs should only resort to \C for asynchronous functionality, 
such as non-blocking I/O, or simple \code{struct} accessors, but never for 
control purposes.%
\footnote{
In \CEU, it is possible to restrict the available \C symbols as a compile-time 
option.
}

@@

\CEU goes one step further and enforces the use of \code{finalize} for 
system calls that deal with pointers representing resources:
%
\begin{itemize}
\item If \CEU \textbf{passes} a pointer to a system call, the pointer 
represents a \textbf{local} resource that requires finalization.
\item If \CEU \textbf{receives} a pointer from a system call return, the 
pointer represents an \textbf{external} resource that requires finalization.
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires 
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
(ln. 5), which is an asynchronous call that transmits the buffer in the 
background.
If the block aborts (ln. 11) before receiving an acknowledge from the 
environment (ln. 9), the local \code{msg} goes out of scope and the external 
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.  
4) returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile without the \code{finalize} construct.

@@

\CEU favors a fine-grained use of trails: it is common to use trails that await 
a single event and terminate.
For this reason, \CEU does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm}: memory for 
blocks in sequence are packed into a \code{union}, while blocks in parallel
are packed into a \code{struct}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at any given time.
A position in the memory may hold different data (with variable sizes) during 
runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with its corresponding 
memory layout.
%
%Each variable is assigned a unique $id$ so that variables with the same name 
%can be distinguished (e.g., \code{a\_1} in \bx:2 corresponding to \code{a} in 
%\ax:3).
%
The \code{do-end} blocks and \code{par/and} in sequence
(ln. \ax:2--4,5--7,8--12) are packed in a \code{union} (ln. \bx:2,3,4--7), 
given that their variables cannot be in scope at the same time, e.g., \code{a} 
and \code{b} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags (ln. \bx:4--7) 
related to the \code{par/and} termination (ln. \ax:8--12), which also reside in 
reusable slots in the static memory.

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences that define 
time units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
We consider that this approach is more familiar to programmers and simplifies 
the reasoning about concurrency.
%
As far as we know, \CEU is the first to encrust this event-triggered notion of 
time in the core of the language, which is a prerequisite for the concurrency 
checks that enable safe shared-memory concurrency.
%
Another unique aspect of \CEU is the distinction between external and internal 
events.
Internal events support stack-based micro reactions within external reactions, 
providing more fine-grained control for intra-reaction execution.

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU goes one step further and performs concurrency checks to detect trails 
that, when reordered, change the observable behavior of the program, i.e., 
trails that actually rely on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the concurrency checks of \CEU.
%
\CEU also guarantees deterministic behavior for timer compositions by adjusting 
for inaccuracies in the system clock.

@@

>>>>

Despite its success and influence, Esterel has a complex semantics that
requires careful static analysis to detect and refuse programs with 
\emph{causality} and \emph{schizophrenia} problems%
%These problems have an extensive coverage in the literature 
~\cite{esterel.constructive,esterel.d7,esterel.d6,esterel.d3,esterel.d5,esterel.d8,esterel.d1,esterel.schizo2}.
%
A complex semantics not only challenges the analysis and compilation of 
programs, but also results in incompatible and non-compliant implementations.
%
Above all, it also affects the programmer's understanding about the code, 
who, ultimately, has to solve the errors when facing corner cases.
%
Another drawback of the Esterel semantics consists of non-deterministic
execution for intra-reaction statements.
Non determinism prevents threads from interacting with stateful system calls safely 
and makes shared-memory concurrency not as straightforward as reading and 
writing to shared variables.
%
%For instance, although Esterel supports orthogonal abortion of 
%threads~\cite{esterel.preemption}, it doesn't offer an effective mechanism to 
%release resources in use.

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU proposes a semantics with fine-grained control for intra-reaction
execution, and is amenable to concurrency checks that improve safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static concurrency checks to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.checks}).
%
\item Safe integration with \C that enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of the proposed semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

TODO: mesmo sem mutual dependency, falar de semantica construtiva e lei da
coerencia

@@

The uniqueness of input events within reactions makes this analysis possible,
otherwise, any two trail segments in parallel could be concurrent, even if they 
react to different input events.
%
Note that the static checks are optional and do not affect the semantics of the
program.
% TODO-no: static analysis also simpler
% highlight that is not only important to have a comptutally-sound algorithm, 
% but that something that programmers can understand

@@

In \CEU, any identifier prefixed with an underscore is passed unchanged to 
the \C compiler that generates the final binary.
Therefore, access to \C is straightforward and syntactically trackable.
%
Similarly to Esterel with the \code{call} primitive, external calls are assumed 
to be instantaneous~\cite{esterel.primer}.
%
This way, programs should only resort to \C for asynchronous functionality, 
such as non-blocking I/O, or simple \code{struct} accessors, but never for 
control purposes.%
\footnote{
In \CEU, it is possible to restrict the available \C symbols as a compile-time 
option.
}

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources:
%
\begin{itemize}
\item If \CEU \textbf{passes} a pointer to a system call, the pointer 
represents a \textbf{local} resource that requires finalization.
\item If \CEU \textbf{receives} a pointer from a system call return, the 
pointer represents an \textbf{external} resource that requires finalization.
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires 
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
(ln. 5), which is an asynchronous call that transmits the buffer in the 
background.
If the block aborts (ln. 11) before receiving an acknowledge from the 
environment (ln. 9), the local \code{msg} goes out of scope and the external 
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.  
4) returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile without the \code{finalize} construct.

@@

\CEU promotes a fine-grained use of trails: it is common to use trails that
await a single event and terminate.
For this reason, \CEU does not allocate per-trail stacks; instead, all data 
resides in fixed memory slots---this is true for the program variables as well 
as for temporary values and runtime flags.
%For instance, the first trail in the guiding example requires temporary slots 
%to hold the locals \code{a} and \code{b}, while the second trail must keep 
%flags to remember which sides of the \code{par/and} have already terminated.
%
Memory for trails in parallel must coexist, while statements in sequence can 
reuse it.
%
Translating this idea to \C is straightforward~\cite{wsn.osm}: memory for 
blocks in sequence are packed into a \code{union}, while blocks in parallel
are packed into a \code{struct}.
%In the example, the code following the loop (identified as \code{...}) reuses 
%all memory from the loop.
%
\CEU reserves a single static block of memory to hold all memory slots, whose 
size is the maximum the program uses at any given time.
A position in the memory may hold different data (with variable sizes) during 
runtime.
%
As an example, Figure~\ref{lst.impl.mem} shows a program with its corresponding 
memory layout.
%
%Each variable is assigned a unique $id$ so that variables with the same name 
%can be distinguished (e.g., \code{a\_1} in \bx:2 corresponding to \code{a} in 
%\ax:3).
%
The \code{do-end} blocks and \code{par/and} in sequence
(ln. \ax:2--4,5--7,8--12) are packed in a \code{union} (ln. \bx:2,3,4--7), 
given that their variables cannot be in scope at the same time, e.g., \code{a} 
and \code{b} can safely share the same memory slot.
%
The example also illustrates the presence of runtime flags (ln. \bx:4--7) 
related to the \code{par/and} termination (ln. \ax:8--12), which also reside in 
reusable slots in the static memory.

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences that define 
time units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
We consider that this approach is more familiar to programmers and simplifies 
the reasoning about concurrency.
%
As far as we know, \CEU is the first to encrust this event-triggered notion of 
time in the core of the language, which is a prerequisite for the concurrency 
checks that enable safe shared-memory concurrency.
%
\CEU also the distinguishes external events from internal events.
Internal events support stack-based micro reactions within external reactions, 
providing more fine-grained control for intra-reaction execution.

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU also performs concurrency checks to detect trails that, when reordered,
change the observable behavior of the program, i.e., trails that actually rely
on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the concurrency checks of \CEU.
%
\CEU also guarantees deterministic behavior for timer compositions by adjusting 
for inaccuracies in the system clock.

@@

TODO: compilação mais sofisticada, tem que lidar com data dependency
